---
title: "College Football WR Performance Study"
author: "Yisong Li"
date: "Monday, February 23, 2015"
output: html_document
---
# Summary
This project will be conducted by a team of students in the Gordon E. Crosby, Jr. MBA Program at the University of Missouri.  The team members are: Edwin Downs, Elie Dvorin, Ryan Frederking, Yisong Li, Phillip Liu, James Stacy, and Kyle Swartz.

## Background
Bill Connelly, a graduate of the Crosby MBA Program, is a well-known sports journalist.  He specializes in analysis and predictability of success for teams as well as players at the collegiate level.  His focus is primarily college football with a little bit of tennis and basketball.  He is an editor for the sports blog site SBNation and Football Outsiders.

## Problem Statement
Bill Connelly has gathered over 10 years of high school and college football data.  This data includes ten years of college football play-by-play data, about 300 games worth of college football charting data, and 10 years of high school recruiting data.  He is looking to employ this data and find more ways to measure position specific efficiency and find ways to tie high school recruiting data to player success at the collegiate level.

## Purpose of study
The purpose of this study is to utilize analytical skills to assist Bill in achieving the results he would like. The team plans to create a logical algorithm(s) that will assist Bill in predicting player success. The team also plans to use knowledge of football and utilize secondary data related to player efficiency at a given position. This will help identify more inputs to produce accurate efficiency measures.

## Scope of study
In this project, the team plans to build a career success rating (CSR) algorithm for college wide receivers based on the true performance rating (TPR) algorithm built during the last project. The team will accept responsibility of collecting players??? recruiting stats from well-known recruiting websites such as ESPN, 247sports, Rivals.com. Finally, team will try to explore the correlations of CSRs, recruiting stats, and other secondary data regard to players, and build a predictive model for the career success level of college wide receivers.

## Limitations
The research and analysis for Bill Connelly will be limited by a couple of factors.  One factor is time constraint.  The team is limited by the duration of the school semester; in turn this will limit the number of positions for which success projects can be made.  Another limitation is the lack of reliable high school recruiting stats. Unlike the professional perspectives, who need to participate the combine, college perspectives normally self-report important physical attributes such as weight, height and 40-yard dash time. These stats could be unreliable because of self-introduced bias and missing data.

# Data processing

## Libraries
```{r Libraries}
suppressMessages(library(data.table))

```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Algorithm of true performance rating (TPR)

```{r TPRAlgorithm}
## Create wrstats function to calculate the wide receivers' yearly stats from 
## pass, player, team, conference tables. The weights arguments set weights for 
## receptions, yards and touchdown, and the f.point and t.limit set the minimum 
## targets and fantasy points required for receivers to be included 
## in the analysis. The default of weights sets to 1 for each reception, 0.1 for 
## each yard, and 3 for each touchdown. You can specify the weights by assigning
## the argument with a numeric vecter with length of 3. The default of minimum 
## targets is 10, and default of minimum fantasy points is 0.

wrstats <- function(pass, player, team, conference, weights = c(1, 0.1, 3), 
                    t.limit = 10, f.limit = 0) {
        wrPlayer <- player[player$Position == "WR",]
        wrPlayer <- merge(wrPlayer, team, by = "Team.Code", all.x = T)
        names(wrPlayer)[ncol(wrPlayer)-1] <- "Team.Name"
        wrPlayer <- merge(wrPlayer, conference, by = "Conference.Code", all.x = T)
        names(wrPlayer)[ncol(wrPlayer)-1] <- "Conference.Name"
        dt <- data.table(pass[,5:9])
        setkey(dt, "Receiver.Player.Code")
        sumDT <- dt[, lapply(.SD,sum), by = key(dt)]
        wrPlayer <- merge(wrPlayer, sumDT, by.x = "Player.Code", 
                          by.y = "Receiver.Player.Code")
        names(wrPlayer)[(ncol(wrPlayer)-3):ncol(wrPlayer)] <-
                c("Targets", "Receptions", "Yards", "Touchdowns")
        wrPlayer$Fantasy.Points <- wrPlayer$Receptions * weights[1] + 
                wrPlayer$Yards * weights[2] + wrPlayer$Touchdowns * weights[3]
        wrPlayer <- subset(wrPlayer, wrPlayer$Fantasy.Points >= f.limit & 
                                   wrPlayer$Targets >= t.limit &
                                   wrPlayer$Subdivision == "FBS")
        output <- wrPlayer[order(-wrPlayer$Fantasy.Points),]
        return(output)
}

## Create ypt function to calculate the adjusted stats because of yards per target from the wide receiver stats data. 
ypt <- function(wr, tier1 = 1, tier2 = 2, adjust1 = 0.025, adjust2 = 0.05){
        Yards.per.Target <- wr$Yards / wr$Targets
        std <- sd(Yards.per.Target)
        mean.ypt <- mean(Yards.per.Target)
        adjust.rate <- vector()
        for (i in 1:length(Yards.per.Target)){
                if (Yards.per.Target[i] > (mean.ypt + tier2 * std)) {
                        adjust.rate[i] = adjust2 
                }
                else if (Yards.per.Target[i] > (mean.ypt + tier1 * std) & 
                                 Yards.per.Target[i] <= (mean.ypt + tier2 * std)){
                        adjust.rate[i] = adjust1
                }
                else if (Yards.per.Target[i] < (mean.ypt - tier1 * std) &
                                 Yards.per.Target[i] >= (mean.ypt - tier2 * std)){
                        adjust.rate[i] = -adjust1
                }
                else if (Yards.per.Target[i] < (mean.ypt - tier2 * std)){
                        adjust.rate[i] = -adjust2
                }
                else{
                        adjust.rate[i] = 0
                }
        }
        Adjusted.YPT.Receptions <- wr$Receptions * adjust.rate
        Adjusted.YPT.Yards <- wr$Yards * adjust.rate
        Adjusted.YPT.Touchdowns <- wr$Touchdowns * adjust.rate
        Adjusted.YPT.Fantasy.Points <- wr$Fantasy.Points * adjust.rate
        output <- data.frame(wr$Player.Code, Yards.per.Target, 
                             Adjusted.YPT.Receptions, Adjusted.YPT.Yards, 
                             Adjusted.YPT.Touchdowns, Adjusted.YPT.Fantasy.Points)
        return(output)
}

## Create thirdDown function to calculate the wide receivers' third down conversion stats from play, pass, player tables.
thirddown <- function(play, pass, wr, weights = c(1, 0.1, 3), third = 0, 
                      forth = 0, adjust = 0.05){
        play$Unique.Code <- paste(play$Game.Code, play$Play.Number, sep = "_")
        pass$Unique.Code <- paste(pass$Game.Code, pass$Play.Number, sep = "_")
        passplay <- merge(pass, play[,c(9,10,15)],by = "Unique.Code")
        passingdown <- (passplay$Down == 3 & passplay$Distance >= third)|
                (passplay$Down == 4 & passplay$Distance >= forth)
        passconversion <- passplay[passplay$X1st.Down == 1 & passingdown, ]
        dt <- data.table(passconversion[,c(6,8,9,10)])
        setkey(dt, "Receiver.Player.Code")
        sumDT <- dt[, lapply(.SD,sum), by = key(dt)]
        setnames(sumDT, 2:4, c("Adjusted.ThirdDown.Receptions", 
                               "Adjusted.ThirdDown.Yards", 
                               "Adjusted.ThirdDown.Touchdowns"))
        output <- data.frame(subset(sumDT, sumDT$Receiver.Player.Code %in% 
                                            wr$Player.Code))
        output$Adjusted.ThirdDown.Receptions <- output[,2] * adjust
        output$Adjusted.ThirdDown.Yards <- output[,3] * adjust
        output$Adjusted.ThirdDown.Touchdowns <- output[,4] * adjust
        output$Adjusted.ThirdDown.Fantasy.Points <- output[,2] * weights[1] +
                output[,3] * weights[2] + output[,4] * weights[2]
        return(output)
}

## Create defensegame function to calculate the pass defense stats of every game. Part of the defense algorithm code, don't need to be called individually.
defensegame <- function(play, pass){
        play$Unique.Code <- paste(play$Game.Code, play$Play.Number, sep = "_")
        pass$Unique.Code <- paste(pass$Game.Code, pass$Play.Number, sep = "_")
        passd <- merge(pass, play[,c(6, 15)],by = "Unique.Code")
        dt <- data.table(passd[,c(2, 7:14)])
        setkeyv(dt, c("Defense.Team.Code","Game.Code"))
        sumDT <- dt[, lapply(.SD,sum), by = key(dt)]
        output <- data.frame(sumDT)
        return(output)
}

## Create defense function to calculate defense adjustment rate of every team for every game with the stat of that game excluded. Part of the defense algorithm code, don't need to be called individually.

defensegameypp <- function(play, pass, g.limit = 1, tier1 = 1, tier2 = 2, 
                           adjust1 = 0.1, adjust2 = 0.2){
        df <- defensegame(play, pass)
        dt <- data.table(df[,c(1,3,5)])
        setkey(dt, "Defense.Team.Code")
        sumDT <- dt[, lapply(.SD,sum), by = key(dt)]
        sumDF <- data.frame(sumDT)
        sumDF$Yards.per.Pass <- sumDF$Yards / sumDF$Attempt
        std <- sd(sumDF$Yards.per.Pass)
        mean.ypp <- mean(sumDF$Yards.per.Pass)
        df <- merge(df, sumDF, by = "Defense.Team.Code", all.x = T)
        df$Adjusted.Attempt <- df$Attempt.y - df$Attempt.x
        df$Adjusted.Yards <- df$Yards.y - df$Yards.x
        df$Adjusted.Yards.per.Pass <- df$Adjusted.Yards / df$Adjusted.Attempt
        df[is.na(df$Adjusted.Yards.per.Pass), 15] <- mean.ypp
        adjust.rate <- vector()
        for (i in 1:nrow(df)){
                if (df[i, 15] > (mean.ypp + tier2 * std)){
                        adjust.rate[i] = -adjust2
                }
                else if (df[i, 15] > (mean.ypp + tier1 * std) & 
                                 df[i, 15] <= (mean.ypp + tier2 * std)){
                        adjust.rate[i] = -adjust1
                }
                else if (df[i, 15] < (mean.ypp - tier1 * std) & 
                                 df[i, 15] >= (mean.ypp - tier2 * std)){
                        adjust.rate[i] = adjust1
                }
                else if (df[i, 15] < (mean.ypp - tier2 * std)){
                        adjust.rate[i] = adjust2
                }
                else{
                        adjust.rate[i] = 0
                }
        }
        df$Adjust.Rate <- adjust.rate
        output <- df[,c(1:2,16)]
        return(output)
}

## Create defense function to calculate the adjusted WR stats according to the defense team
defense <- function(wr, play, pass, weights = c(1, 0.1, 3), g.limit = 1, tier1 = 1, tier2 = 2, 
                    adjust1 = 0.1, adjust2 = 0.2) {
        df <- defensegameypp(play, pass, g.limit, tier1, tier2, adjust1, adjust2)
        passd <- merge(pass, play[,c(1, 2, 6)],by = c("Game.Code", "Play.Number"))
        passdadj <- merge(passd, df, by = c("Defense.Team.Code", "Game.Code"), all.x = T)
        passdadj$Adjusted.Defense.Receptions <- passdadj$Completion * passdadj$Adjust.Rate
        passdadj$Adjusted.Defense.Yards <- passdadj$Yards * passdadj$Adjust.Rate
        passdadj$Adjusted.Defense.Touchdown <- passdadj$Touchdown * passdadj$Adjust.Rate
        dt <- data.table(passdadj[,c(6, 15:17)])
        setkey(dt, "Receiver.Player.Code")
        sumDT <- dt[, lapply(.SD,sum), by = key(dt)]
        output <- data.frame(sumDT)
        output$Adjusted.Defense.Fantasy.Point <- output[,2] * weights[1] + 
                output[,3] * weights[2] + output[,4] * weights[3]
        return(output)
}

## Create the TPR function to call all the algorithm, and output the final WR stats. No need to change the parameters right now, default should be fine.
TPR <- function(play, pass, player, team, conference, weights = c(1, 0.1, 3), 
                t.limit = 10, f.limit = 0, g.limit = 1, ypt.tier1 = 1, 
                ypt.tier2 = 2, ypt.adjust1 = 0.025, ypt.adjust2 = 0.05, 
                third = 0, forth = 0, t.adjust = 0.05, d.tier1 = 1, d.tier2 = 2, 
                d.adjust1 = 0.1, d.adjust2 = 0.2){
        wr <- wrstats(pass, player, team, conference, weights, t.limit, f.limit)
        yptadj <- ypt(wr, ypt.tier1, ypt.tier2, ypt.adjust1, ypt.adjust2)
        tadj <- thirddown(play, pass, wr, weights, third, forth, t.adjust)
        dadj <- defense(wr, play, pass, weights, g.limit, d.tier1, d.tier2, 
                        d.adjust1, d.adjust2)
        final <- merge(wr,yptadj, by.x = "Player.Code", by.y = "wr.Player.Code", all.x = T)
        final <- merge(final, tadj, by.x = "Player.Code", by.y = "Receiver.Player.Code", all.x = T)
        final <- merge(final, dadj, by.x = "Player.Code", by.y = "Receiver.Player.Code", all.x = T)
        final$TPR <- final$Fantasy.Points + final$Adjusted.YPT.Fantasy.Points + 
                final$Adjusted.ThirdDown.Fantasy.Points + 
                final$Adjusted.Defense.Fantasy.Point
        final <- final[order(-final$TPR),]
        rownames(final) <- NULL
        return(final)
}
```
## Algorithm of career success rating (CSR)

```{r CSRAlgorithm}

```
## Set the constants and working directory
```{r SetConstants}
## Set the constant used for this program. The STARTYEAR is the start year of
## the available WR data, the ENDYEAR is the end year of the available WR data.
## The DATAPATH is the folder paths for each year's data. I recommend to put all
## the data folders under a "data" folder under the working directory. 
STARTYEAR <- 2005; ENDYEAR <- 2014
YEAR <- STARTYEAR:ENDYEAR
DATAPATH <- c("./data/collegefootballdata.org-2005-1.2.0/",
              "./data/collegefootballdata.org-2006-1.5.0/",
              "./data/collegefootballdata.org-2007-1.5.0/", 
              "./data/collegefootballdata.org-2008-1.5.0/", 
              "./data/collegefootballdata.org-2009-1.5.0/", 
              "./data/collegefootballdata.org-2010-1.5.0/", 
              "./data/collegefootballdata.org-2011-1.5.0/", 
              "./data/collegefootballdata.org-2012/", 
              "./data/collegefootballdata.org-2013/", 
              "./data/collegefootballdata.org-2014/")

## WD is the path of your working directory, where you put the "data" folder and
## the "Function script.R" file. Please change the path according the setup of 
## your computer. Be careful about the format. If you are not sure what is the
## right format, you can use the file.path function. Be sure to use ?file.path
## call the help page and read how to use it.
WDPATH <- "D:/Baidu Cloud/Crosby MBA Program 2015 Spring/BA 8600 - Consulting/R programing/"

## Set the working directory according the "WD" constant
setwd(WDPATH)
```

## Read in the datasets and calculate the TPR and CSR
```{r ReadData, cache=TRUE}
## Loop through all the data by years and automatically generate TPR file for
## each year under the folder "TPR" that created earlier.
play <- list()
pass <- list()
player <- list()
team <- list()
conference <- list()
for (i in 1:length(DATAPATH)){
        play[[paste("play", YEAR[i], sep = "")]] <- 
                read.csv(paste(DATAPATH[i], "play.csv", sep = ""))
        pass[[paste("pass", YEAR[i], sep = "")]] <- 
                read.csv(paste(DATAPATH[i], "pass.csv", sep = ""))
        player[[paste("player", YEAR[i], sep = "")]] <- 
                read.csv(paste(DATAPATH[i], "player.csv", sep = ""))
        team[[paste("team", YEAR[i], sep = "")]] <- 
                read.csv(paste(DATAPATH[i], "team.csv", sep = ""))
        conference[[paste("conference", YEAR[i], sep = "")]] <- 
                read.csv(paste(DATAPATH[i], "conference.csv", sep = ""))
}
```
## Calculate the TPRs and CSRs
```{r TPRCalculation, cache=TRUE}
TPRdata <- list()
for (i in 1:length(DATAPATH)){
        TPRdata[[paste("WR", YEAR[i], sep = "")]] <- 
                TPR(play[[i]], pass[[i]], player[[i]], team[[i]], conference[[i]])
}
str(TPRdata)
```
# Analysis

# Results


